Linear response theory for core set identification

### 2012-05-15 writedown/linear.response/langevin.tex
add file for langevin dynamics

\newcommand{\mh}{\mathcal H}
\newcommand{\ml}{\mathcal L}
\newcommand{\mt}{\mathcal T}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}

@@

example. Here we assume the system is governed by equation:

@@

  \dot{\vect q}_i & = \frac{\partial\mh}{\partial \vect p_i}
  + \vect C_i(\vect q,\vect p) F_e(t)\\\label{eqn:gov-2}
  \dot{\vect p}_i & =- \frac{\partial\mh}{\partial \vect q_i}
  + \vect D_i(\vect q,\vect p) F_e(t)

@@

  \sum_{i=1}^{3N}
  \frac{\partial \vect C_i}{\partial \vect q_i} +
  \frac{\partial \vect D_i}{\partial \vect p_i}
  \bigg] = 0

@@

q_i, \vect p_i\}$) is subject to the Liouville equation:

@@

  \frac{\partial}{\partial t} f(\vect x, t) + i\ml f(\vect x, t) = 0

@@

where $\ml$ is the Liouville operator given by $i\ml = \dot{\vect
  x}\cdot\nabla_{\vect x}$.  We assume that the perturbation to the

@@

Liouville operator can be viewed as perturbation from the equilibrium

@@

respectively. Namely

@@

and since the Eqn.~\eqref{eqn:gov-1} and \eqref{eqn:gov-2} imply that
$\dot{\vect x} = \dot{\vect x}_0 + \Delta \dot{\vect x}(t)$

@@

  i\ml =
  (\dot{\vect x}_0 + \Delta \dot{\vect x}(t))\cdot\nabla_{\vect x}
  = i\ml_0 + i\Delta\ml(t)

@@

Where the invariant measure satisfies $i\ml_0f_0(\vect x) = 0$. Also
$f_0(\vect x)$ has the form $f_0(\mh(\vect x))$ (?). The Liouville equation
for the perturbed system is therefore:

@@

  [f_0(\mh(\vect x)) + \Delta  f(\vect x, t)]
  +
  [i\ml_0 + i\Delta\ml(t)]
  [f_0(\mh(\vect x)) + \Delta  f(\vect x, t)] = 0

@@

term $i\Delta\ml(t)  \Delta  f(\vect x, t)$ in the Liouville equation
can be neglected. Noticing $i\ml_0 f_0 = 0$, we have the Liouville equation
for the leading order of perturbation:

@@

  + i\ml_0

@@

  - i\Delta\ml(t) f_0(\mh(\vect x))

@@

The operator on l.h.s. is the same as the unperturbed Liouville equation.

@@

  - \int_0^t \dd s \:
  e^{-i\ml_0(t-s)}
  i\Delta\ml(s)f_0(\mh(\vect x))

@@

  -\sum_{i=1}^{3N}
  \vect D_i(\vect x) \frac{\partial\mh}{\partial \vect p_i} +
  \vect C_i(\vect x) \frac{\partial\mh}{\partial \vect q_i}   

@@

  i\Delta\ml(s)f_0(\mh(\vect x)) =
  -\frac{\partial f_0(\mh)}{\partial\mh} j(\vect x)\,F_e(s)
For the canonical invariant measure, this is equivalent to
  i\Delta\ml(s)f_0(\mh(\vect x)) =
  \beta f_0(\mh(\vect x)) j(\vect x)F_e(s)

@@

  \mathcal A(t)

@@

  \int\dd \vect x A(\vect x)f(\vect x, t)  \\\nonumber

@@

  \int\dd \vect x\, A(\vect x)\,
  [f_0(\mh(\vect x)) + \Delta f(\vect x, t)] \\\nonumber

@@

  \langle A\rangle_0

@@

  \int \dd \vect x
  A(\vect x)

@@

  e^{-i\ml_0(t-s)}
  f_0(\mh(\vect x)) j(\vect x)F_e(s)  \\\nonumber

@@

  A(\vect x)
  e^{-i\ml_0(t-s)}
  f_0(\mh(\vect x)) j(\vect x)  \\

@@

  f_0(\mh(\vect x))
  A(\vect x)
  e^{-i\ml_0(t-s)}
  j(\vect x)  
Notice that $e^{i\ml_0(t-s)}A(\vect x) = A(\phi_{t-s}(\vect x))$,
where $\phi_t(\vect x)$ is the flow mapping of the \emph{unperturbed}
Hamiltonian system. We have
  \mathcal A(t)
  \langle A\rangle_0

@@

@@

  f_0(\mh(\vect x))
  A(\phi_{t-s}(\vect x))
  j(\vect x)  

@@

  A(t-s)\,

@@

@@

where $\langle A(t-s)\,j(0)\,\rangle_0$ is the \emph{equilibrium time
  correlation} between the observation $A$ at time $t-s$ and the

@@

The core set plays a very important role in the MSM theory. One way to
identify core set is to consider the ``stability'' of the invariant
measure $f_0$ under some mapping $\mt_\alpha$ of time step $\alpha$:
\begin{align}
  C_\alpha = \big\{
  \vect x \,\vert\, \mt_\alpha f_0(\vect x) > f_0(\vect x)
  \big\}
\end{align}
Generally, it is impossible and not necessary to consider the core set
in the $3N$ dimensional configuration space. For example, a
macromolecule solving in water, the DOFs of water is not of interest
at all. Therefore, we consider some phase space compressing function
$\vect y = \vect y(\vect x)$, with $\vect y \in \Omega \subset \mathbb
R^M,\ M\ll 3N$. We call $\Omega$ the observation space.  The projected
invariant measure on the observation space is therefore
\begin{align}
  \mathcal P f_0 (\vect y)
  =
  \int
  \dd \vect x\,
  f_0(\vect x) \,
  \delta (\vect y - \vect y(\vect x)) 
\end{align}
The core set on the observation space is
\begin{align}
  \widetilde
  C_\alpha = \big\{
  \vect y \,\vert\,
  \mathcal P\mt_\alpha f_0(\vect y) > \mathcal P f_0(\vect y)
  \big\}
\end{align}
% with $\widetilde \mt_\alpha$ defined by
% $\widetilde\mt_\alpha = \mathcal P\circ\mt_\alpha$, namely
% \begin{align}
%   \widetilde\mt_\alpha\, p_0(\vect y)
%   =
%   \mathcal P\circ\mt_\alpha f_0 (\vect y)
%   =
%   \int
%   \dd \vect x\,
%   [\mt_\alpha f_0](\vect x) \,
%   \delta (\vect y - \vect y(\vect x)) 
% \end{align}
For convenience, we define $\Delta \mt_\alpha =
\mt_\alpha - \mathcal I$, where $\mathcal I$ is the
identity. If $[\mathcal P\circ\Delta \mt_\alpha f_0](\vect y) > 0$, then
$\vect y$ is in the core set $\widetilde C_\alpha $.
\begin{align}\nonumber
  [\mathcal P\circ\Delta \mt_\alpha f_0](\vect y)
  &=
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  \delta (\vect y - \vect y(\vect x))  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  K_{\sigma} (\vect y - \vect y(\vect x))  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  K_{\sigma, \vect y} (\vect x)  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  f_0(\vect x) \,
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y} (\vect x)\\
  &=
  \lim_{\sigma \rightarrow 0}\,
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_0
\end{align}
where $[\Delta \mt_\alpha]^\dagger$ is the adjoint operator of $\Delta
\mt_\alpha$, which is actually $\Delta \mt_{-\alpha}$. The delta
function is replaced by the limit of a mollifier $K_\sigma(\vect y)$,
which is infinitely smooth, compact supported of size $\sigma$,
bounded and $\int
K_\sigma(\vect y)\dd \vect y = 1$.  The physical meaning is that we
can consider the expectation value of the stability of the mollifier
$K_{\sigma, \vect y}$ instead of the stability of the invariant measure.
This provides the convenience of using the linear response theory showed
in the previous section. By Eqn.~\eqref{eqn:lr},
\begin{align}\nonumber
  [\mathcal P\circ\Delta \mt_\alpha f_t](\vect y)
  &=
  \lim_{\sigma \rightarrow 0}\,
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_t \\
  &=
  \lim_{\sigma \rightarrow 0}\,
  \Big\{
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_0  -
  \beta
  \int_0^t\dd s\,
  F_e(s)
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}(t-s)\,
  j(0)\,
  \rangle_0
  \Big\}
\end{align}
Therefore, the core set at time $t$ is defined by
\begin{align}
  \widetilde C_\alpha^t =
  \Big\{
  \vect y
  \,\Big\vert\,
  [\mathcal P\circ\Delta \mt_\alpha f_t](\vect y) > 0
  \Big\}
\end{align}

>>>>

\newcommand{\mh}{\mathcal H}
\newcommand{\ml}{\mathcal L}
\newcommand{\mt}{\mathcal T}
\newcommand{\fwg}{{\mathcal A}}
\newcommand{\bwg}{{\mathcal B}}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}

@@

example. Here we assume the system is governed by the
perturbed Langevin equation:

@@

  \dot{\vect q} & = \frac{\partial\mh}{\partial \vect p}
  + \vect C(\vect q,\vect p) F_e(t)\\\label{eqn:gov-2}
  \dot{\vect p} & =- \frac{\partial\mh}{\partial \vect q}
  + \vect D(\vect q,\vect p) F_e(t)
  - \gamma\vect p
  + \sigma\dot{\vect W}

@@

  \nabla_{\vect q}\cdot\vect C + \nabla_{\vect p}\cdot\vect D = 0

@@

q, \vect p\}$) is subject to the Fokker-Planck equation:

@@

  \frac{\partial}{\partial t} f(\vect x, t) - \fwg f(\vect x, t) = 0

@@

where $\fwg$ is the infiniesimal generator gieven by
  \fwg =
  \frac{\sigma^2}2\Delta_{\vect p}
  - \Big(
  \frac{\partial \mh}{\partial\vect p} + \vect C F_e(t)
  \Big)\cdot\nabla_{\vect q}
  - \Big(
  -\frac{\partial \mh}{\partial\vect q} +
  \vect D F_e(t) - \gamma\vect p
  \Big)\cdot\nabla_{\vect p}
  + 3N\gamma
The equilibrium and unperturbed system
is governed by
the standard Langevin equation, i.e. with
vanishing input perturbation $F_e(t) = 0$ in
Eqn.~\eqref{eqn:gov-1} and \eqref{eqn:gov-2}.
We assume that the perturbation to the

@@

infiniesimal generator can be viewed as perturbation from the equilibrium

@@

respectively. We have

@@

and 

@@

  \fwg = \fwg_0 + \Delta\fwg(t)

@@

Where the invariant measure satisfies $\fwg_0f_0(\vect x) = 0$,
which is nothing but the Boltzmann distribution
  f_0(\vect x) \propto e^{-\beta\mh(\vect x)}
The Fokker-Planck equation for the perturbation is therefore

@@

  [f_0(\vect x) + \Delta  f(\vect x, t)]
  [\,\fwg_0 + \Delta\fwg(t)\,]
  [\,f_0(\vect x) + \Delta  f(\vect x, t)\,] = 0

@@

term $\Delta\fwg(t)  \Delta  f(\vect x, t)$ in the Fokker-Planck equation
can be neglected. Noticing $\fwg_0 f_0(\vect x) = 0$, we have the
Fokker-Planck
for the leading order of the perturbation:

@@

  - \fwg_0

@@

  \Delta\fwg(t) f_0(\vect x)

@@

The operator on l.h.s. is the same as the unperturbed Fokker-Planck equation.

@@

  \int_0^t \dd s \:
  e^{(t-s)\fwg_0}
  \Delta\fwg(s)f_0(\vect x)

@@

  -\vect C\cdot\nabla_{\vect q}\mh 
  -\vect D\cdot\nabla_{\vect p}\mh

@@

  \Delta\fwg(s)f_0(\vect x) =
  -\beta f_0(\vect x)\, j(\vect x) F_e(s)

@@

  \mathcal O(t)

@@

  \int\dd \vect x \:O(\vect x)f(\vect x, t)  \\\nonumber

@@

  \int\dd \vect x\, O(\vect x)\,
  [f_0(\vect x) + \Delta f(\vect x, t)] \\\nonumber

@@

  \langle O\rangle_0

@@

  \int \dd \vect x\:
  O(\vect x)

@@

  e^{(t-s)\fwg_0}
  f_0(\vect x)\, j(\vect x)F_e(s)  \\\nonumber

@@

  O(\vect x)\,
  e^{(t-s)\fwg_0}
  f_0(\vect x) j(\vect x)  \\

@@

  f_0(\vect x) j(\vect x)\,
  e^{(t-s)\bwg_0}
  O(\vect x)\\
  \langle O\rangle_0

@@

  F_e(t-s)

@@

  f_0(\vect x) j(\vect x)\,
  e^{s\bwg_0}
  O(\vect x)
Where $\bwg_0$ is the infiniesimal generator of the \emph{unperturbed} backward
Kolmogorov equation, which is the adjoint operator of $\fwg_0$.
We have
  e^{s\bwg_0} O(\vect x) = \mathbb E_{\vect x} [O(\vect X_s)]
where $\vect X_t$ is the trajectory of the unperturbed Langevin equation
starting at $\vect x$.
Therefore, we have
% Notice that $e^{i\ml_0(t-s)}O(\vect x) = O(\phi_{t-s}(\vect x))$,
% where $\phi_t(\vect x)$ is the flow mapping of the \emph{unperturbed}
% Hamiltonian system. We have
  \mathcal O(t)
  \langle O\rangle_0
  F_e(t-s)
  f_0(\vect x)
  j(\vect x)\,
  \mathbb E_{\vect x} [O(\vect X_s)]

@@

@@

  \mathbb E_{\vect x} [O(\vect X_s)]

@@

where $  \langle
  \mathbb E_{\vect x} [O(\vect X_s)]
  \rangle_0$ is the \emph{equilibrium time
  correlation} between the observation $O$ at time $s$ and the

@@

The core set plays a very important role in the MSM theory. One way to
identify core set is to consider the ``stability'' of the invariant
measure $f_0$ under some mapping $\mt_\alpha$ of time step $\alpha$:
\begin{align}
  C_\alpha = \big\{
  \vect x \,\vert\, \mt_\alpha f_0(\vect x) > f_0(\vect x)
  \big\}
\end{align}
Generally, it is impossible and not necessary to consider the core set
in the $3N$ dimensional configuration space. For example, a
macromolecule solving in water, the DOFs of water is not of interest
at all. Therefore, we consider some phase space compressing function
$\vect y = \vect y(\vect x)$, with $\vect y \in \Omega \subset \mathbb
R^M,\ M\ll 3N$. We call $\Omega$ the observation space.  The projected
invariant measure on the observation space is therefore
\begin{align}
  \mathcal P f_0 (\vect y)
  =
  \int
  \dd \vect x\,
  f_0(\vect x) \,
  \delta (\vect y - \vect y(\vect x)) 
\end{align}
The core set on the observation space is
\begin{align}
  \widetilde
  C_\alpha = \big\{
  \vect y \,\vert\,
  \mathcal P\mt_\alpha f_0(\vect y) > \mathcal P f_0(\vect y)
  \big\}
\end{align}
% with $\widetilde \mt_\alpha$ defined by
% $\widetilde\mt_\alpha = \mathcal P\circ\mt_\alpha$, namely
% \begin{align}
%   \widetilde\mt_\alpha\, p_0(\vect y)
%   =
%   \mathcal P\circ\mt_\alpha f_0 (\vect y)
%   =
%   \int
%   \dd \vect x\,
%   [\mt_\alpha f_0](\vect x) \,
%   \delta (\vect y - \vect y(\vect x)) 
% \end{align}
For convenience, we define $\Delta \mt_\alpha =
\mt_\alpha - \mathcal I$, where $\mathcal I$ is the
identity. If $[\mathcal P\circ\Delta \mt_\alpha f_0](\vect y) > 0$, then
$\vect y$ is in the core set $\widetilde C_\alpha $.
\begin{align}\nonumber
  [\mathcal P\circ\Delta \mt_\alpha f_0](\vect y)
  &=
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  \delta (\vect y - \vect y(\vect x))  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  K_{\sigma} (\vect y - \vect y(\vect x))  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  \Delta \mt_\alpha f_0(\vect x) \,
  K_{\sigma, \vect y} (\vect x)  \\\nonumber
  &=
  \lim_{\sigma \rightarrow 0}
  \int
  \dd \vect x\,
  f_0(\vect x) \,
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y} (\vect x)\\
  &=
  \lim_{\sigma \rightarrow 0}\,
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_0
\end{align}
where $[\Delta \mt_\alpha]^\dagger$ is the adjoint operator of $\Delta
\mt_\alpha$, which is actually $\Delta \mt_{-\alpha}$. The delta
function is replaced by the limit of a mollifier $K_\sigma(\vect y)$,
which is infinitely smooth, compact supported of size $\sigma$,
bounded and $\int
K_\sigma(\vect y)\dd \vect y = 1$.  The physical meaning is that we
can consider the expectation value of the stability of the mollifier
$K_{\sigma, \vect y}$ instead of the stability of the invariant measure.
This provides the convenience of using the linear response theory showed
in the previous section. By Eqn.~\eqref{eqn:lr},
\begin{align}\nonumber
  [\mathcal P\circ\Delta \mt_\alpha f_t](\vect y)
  &=
  \lim_{\sigma \rightarrow 0}\,
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_t \\
  &=
  \lim_{\sigma \rightarrow 0}\,
  \Big\{
  \langle
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}
  \rangle_0  -
  \beta
  \int_0^t\dd s\,
  F_e(t-s)
  \langle
  j(0)\,
  \mathbb E_{\vect x}\{
  [\Delta \mt_\alpha]^\dagger K_{\sigma, \vect y}(s)\,
  \}
  \rangle_0
  \Big\}
\end{align}
Therefore, the core set at time $t$ is defined by
\begin{align}
  \widetilde C_\alpha^t =
  \Big\{
  \vect y
  \,\Big\vert\,
  [\mathcal P\circ\Delta \mt_\alpha f_t](\vect y) > 0
  \Big\}
\end{align}
